---
title: "Yellow-naped amazon xcorr analyses and PCA"
author: "Molly Dupin"
date: "7/18/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, read in data}
setwd("~/Desktop/R/MS_thesis/Files_for_final_analysis/Vocal_data")
getwd()

X <- c("warbleR", "Rraven", "parallel", "devtools", "dplyr", "ggfortify", "ggplot2", "pbapply", "plyr", "tidyverse", "mclust", "aricode", "ks", "overlap", "data.table", "scales", "viridis", "egg", "ggplotify", "grid", "effsize", "userfriendlyscience", "car", "rosetta")
invisible(lapply(X, library, character.only = TRUE))

# Read in extended selection table
sels <- readRDS("Yellow_naped_vocal_data_EST.RDS")
View(sels)

# Check the available number of cores
parallel::detectCores()

# set the cores to use less than 2 of the total
cores <- parallel::detectCores() - 2 

```

We use spectrographic cross-correlation (SPCC) to estimate pairwise dissimilarity across all calls using the cross_correlation function. We created multi-dimensional space plots to visualize the data.

```{r run SPCC for all calls and visualize}
# Run the spectrographic cross-correlation using optimal parameters
xc_mat <- xcorr(X = sels, wl = 512, ovlp = 85, bp = c(0.5, 2.5), parallel = cores, pb = TRUE, cor.mat = TRUE, na.rm = TRUE)
str(xc_mat)
View(xc_mat)

# Save the generated matrix as RDS file
saveRDS(xc_mat, "SPCC_YNA_vocal_data.RDS")

# Make the matrix an object to use for plotting
xc_mat <- readRDS("SPCC_YNA_vocal_data.RDS")

# Look at calls using multidimensional scaling (MDS) to visualize SPCC acoustic space
?stats::cmdscale

# Create a distance matrix for plotting the points in acoustic space
xc_dist <- stats::as.dist(1-xc_mat, diag = TRUE, upper = FALSE) 
class(xc_dist)

# Use k=2 to view in 2 dimensions using the distance matrix points
mds_res <- stats::cmdscale(d = xc_dist, k = 2, add = TRUE)

# Extract site names from xc_mat dimension names
nms <- dimnames(xc_mat)[[1]]
site <- sapply(1:length(nms), function(x){
  strsplit(nms[x], split = "_")[[1]][5]
})

# Create a data frame out of the 2 dimensions, first 2 columns, for plotting calls by type
df <- data.frame(X = mds_res$points[, 1], Y = mds_res$points[, 2], Call_Type = sels$Call_Type)
str(df)

# Make a color palette object with different colors for each call type Be careful about using predetermined color palettes, as some colors may be repeated (if number of colors in palette are fewer than number of call types) or too similar to visually distinguish once plotted
cbPalette <- c("red", "blue", "gold", "purple4", "orange", "green4", "lightpink", "deepskyblue", "magenta", "lightgoldenrod", "grey", "brown", "black", "red4", "orchid4", "tan1", "hotpink4", "indianred", "orchid", "lightseagreen", "lightseablue", "darkolivegreen1", "rosybrown4", "darkolivegreen4", "tan4", "honeydew4")

# Set the aesthetics for the graph background/layout
theme_AcousticSpace <- function(){
  theme(panel.background = element_rect(fill = "white", color = "white"), 
        plot.background = element_rect("white", color = "white"),
        panel.grid.major = element_line(colour = "white"),
        panel.grid.minor = element_line(colour = "white"),
        axis.line = element_line(size = 0.8, colour = "black"),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 25),
        axis.text = element_text(size = 15))
}

# Points plotted using hulls will have polygons around each call type, skip this step to create plots without polygons
# hulls <- plyr::ddply(df, "Call_Type", function(x){x[chull(x$X, x$Y), ]})

ggplot(data = df, 
       aes(x = X, y = Y, color = Call_Type, fill = Call_Type)) +
  geom_point(size = 7, shape = 21) +
  # geom_polygon(data = hulls, aes(x = X, y = Y, fill = Call_Type, color = Call_Type), alpha = 0.2, size = 0.2) + # Leave this commented out if you are creating plots without polygons
  scale_color_manual(values = cbPalette) + 
  scale_fill_manual(values = cbPalette) +
  xlab("MDS Dimension 1") + 
  ylab("MDS Dimension 2") +
  theme_AcousticSpace() +
  theme(legend.position = "right") +

# Export a .jpg file of the plot
jpeg("MDS_all_vocal_data.jpeg", units = NULL, width = 10, height = 11, res = 300)

```

In order to better visualize the data we generated MDS plots for each country sampled. This process creates separate cross correlation matrices for each country, they can be saved as individual .RDS files to access in the future for quick country-level plotting without rerunning the xcorr function


```{r create MDS plots by countries}

# Note: instead of rewriting the same code for every country, we changed the value within the grep function to whatever country we wanted to plot (Mexico, Guatemala, Honduras, Nicaragua, Costa Rica). We also changed the name of objects created to include a country 'tag' (ie, Mex, Guat, Hond, Nica, Costa) to avoid confusion with the previously made matrix for all calls

# grep out specific group (can be region, country, or any categorical variable)
xc_mat_Mex <- xcorr(X = sels[grep("Mexico", sels$Country), ], wl = 512, ovlp = 85, bp = c(0.5, 2.5), parallel = cores, pb = TRUE, cor.mat = TRUE, na.rm = TRUE)
str(xc_mat_Mex)
View(xc_mat_Mex)
# saveRDS(xc_mat_Mex, file = "SPCC_Mexico_calls.RDS")

# Create the distance matrix
xc_dist_Mex <- stats::as.dist(1-xc_mat_Mex, diag = TRUE, upper = FALSE)
class(xc_dist_Mex)

# Use only 2 dimensions b/c just visualizing
mds_res_Mex <- stats::cmdscale(d = xc_dist_Mex, k = 2, add = TRUE)

# Create a data frame to use for plotting
# first extract site names from xc_mat dimension names
nms_Mex <- dimnames(xc_mat_Mex)[[1]]
site <- sapply(1:length(nms), function(x){
  strsplit(nms[x], split = "_")[[1]][5]
})

temp_Mex <- filter(sels, grepl("Mexico", sels$Country))
View(temp_Mex)
temp_Call_Type <- temp_Mex$Call_Type
View(temp_Call_Type)

df_Mex <- data.frame(X = mds_res_Mex$points[, 1], Y = mds_res_Mex$points[, 2], Call_Type = temp_Mex$Call_Type)
str(df)

hulls <- plyr::ddply(df_Mex, "Call_Type", function(x){x[chull(x$X, x$Y), ]})

theme_AcousticSpace <- function(){
  theme(panel.background = element_rect(fill = "white"), plot.background = element_rect(fill = "white"),
        panel.grid.major = element_line(size = 3, colour = "white"),
        panel.grid.minor = element_line(size = 0.75, colour = "white"),
        axis.line = element_line(size = 0.45, colour = "black"),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 12))
}

# make a color palette object with number of different colors equivalent to number of call types
# Note: you may get an error code if number of colors listed does not match number of call types. Instead of deleting them and readding them constantly, just comment out what is not needed

cbPalette <- c("cornflowerblue", "coral4", "chartreuse1", "darkorchid4") # "coral", "green4", "cadetblue1", "darkslateblue", "black", "darkolivegreen", "darkmagenta", "azure4", "firebrick2", "blue", "deeppink", "grey50", "hotpink4", "khaki1", "springgreen1", "cyan", "plum1", "steelblue", "wheat4"

# Note: shape can be another useful way to distinguish call types if necessary
jpeg("Mexico_SPCC_MDS_Jul19.jpeg", units = NULL, width = 10, height = 10, res = 300)
ggplot(data = df_Mex, 
       aes(x = X, y = Y, color = temp_Call_Type, fill = temp_Call_Type)) +
  geom_point(size = 3) + # shape = temp_Call_Type
  # geom_polygon(data = hulls, aes(x = X, y = Y, fill = temp_Call_Type, color = temp_Call_Type), alpha = 0.2, size = 0.2) +
  ggtitle("YNA vocal data clustered by visually classified call types from Mexico") +
  scale_color_manual(values = cbPalette) + 
  scale_fill_manual(values = cbPalette) +
  xlab("MDS Dimension 1") + 
  ylab("MDS Dimension 2") +
  theme_AcousticSpace() +
  theme(legend.position = "top") 
dev.off()
```

Kernel plots were then produced to show the distribution of each call using contour lines, with closer lines representing a higher density of calls. The smallest circles indicate high concentrations of calls in one area of acoustic space.

```{r kernel plots for each country}

# create path object
path <- "~/Desktop/R/MS_thesis/Files_for_final_analysis/Vocal_data"

# Read in extended selection table 
ccs_est <- readRDS("Yellow_naped_vocal_data_EST.RDS")
glimpse(ccs_est)

# Check available cores
parallel::detectCores()

# Set to use 2 less cores than available
cores <- parallel::detectCores() - 2

# Read in the cross correlation matrix
xc_mat <- readRDS(file.path(path, "SPCC_YNA_vocal_data.RDS"))
str(xc_mat)

# Create distance matrix
xc_mat_dist <- stats::as.dist(1 - xc_mat, diag = TRUE, upper = TRUE)
str(xc_mat_dist)

# Perform MDS to get 2 dimensions; k may need to be changed depending on visible patterns in low dimensional acoustic space

# Note: The higher k is set, the lower maxit should be set
iso <- invisible(MASS::isoMDS(xc_mat_dist, k = 4, maxit = 1000, trace = FALSE))
str(iso)
saveRDS(iso, "YNA_kernel_isoMDS.RDS")

# Read in iso matrices here if you've already made it
iso <- readRDS("YNA_kernel_isoMDS.RDS") 

# Combine the call matrix with information from metadata sheet for the acoustic space plot
meta <- readRDS("Yellow_naped_amazon_vocal_metadata.RDS")
kern_df <- meta %>%
  dplyr::select(sound.files, Year, Country, Region, site, Call_Type, site_year) %>% 
dplyr::mutate(
  X = iso$points[,1], Y = iso$points[,2])
glimpse(kern_df)

# This can be saved as an RDS or csv when finished
saveRDS(kern_df, file.path(path, "kernel_iso_with_metadata.RDS"))

# Read in the RDS file for the isoMDS if its already been generated
kern_df <- readRDS("kernel_iso_with_metadata.RDS")

regions <- kern_df %>%
  pull(Region) %>%
  unique()

Country <- kern_df %>%
  pull(Country) %>%
  unique()

View(kern_df)

# Make a color palette
cols <- c("red", "blue", "gold", "purple4", "orange", "green4", "lightpink", "deepskyblue", "magenta", "lightgoldenrod", "grey", "brown", "black", "red4", "orchid4", "tan1", "hotpink4", "indianred", "orchid", "lightseagreen", "lightseablue", "darkolivegreen1", "rosybrown4", "darkolivegreen4", "tan4", "honeydew4")

x <- c(4) # change number based on the country you want to plot: 1 (Costa Rica), 2 (Nicaragua), 3 (Mexico), 4 (Honduras), 5 (Guatemala)
y <- 1
kern_df_tmp <- kern_df %>%
  filter(Country == country[x]) %>%
  droplevels() 

# Note: change .jpeg title to reflect country being plotted
ggplot(kern_df_tmp, aes(x = X, y = Y, color = Call_Type)) +
  geom_density2d() +
  scale_color_manual(values = cols) +
  xlab("MDS Dim 1") +
  ylab("MDS Dim 2") +
  theme(panel.background = element_rect(fill = "white", color = "white"), 
        plot.background = element_rect("white", color = "white"),
        panel.grid.major = element_line(colour = "white"),
        panel.grid.minor = element_line(colour = "white"),
        axis.line = element_line(size = 0.8, colour = "black"),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 25),
        axis.text = element_text(size = 15),
        legend.position = "right") +
jpeg("Kernel_plot_YNA.jpeg", units = NULL, width = 10, height = 11, res = 300)

```

We used principal component analysis on spectrographic features as an alternative approach to measure similarity between call types across the range. We plotted the first 2 PC's for each country to visualize results. We then performed a one-way ANOVA to test the means, and conduct a post-hoc test using posthocTGH().

```{r perform PCA, conduct ANOVA and post-hoc testing}

# Read in extended selection table 
sels <- readRDS("Yellow_naped_vocal_data_EST.RDS")
variant <- sels$Call_Type

# To conduct PCA on calls by country, use the filter function to pull out calls from desired country
# temp <- filter(sels, grepl("Mexico", sels$Country))
# variant <- temp$Call_Type
# str(temp)

# Measure acoustic parameters of all calls
param <- specan(sels, bp = c(0.5, 2.5), ovlp = 85, ff.method = "tuneR", wn = "hanning") # [grep("Mexico", sels$Country),]
str(param)
View(param)

# Remove columns that would confuse R
param$selec <- NULL
# temp$selec <- NULL
str(param)

# Conduct PCA
pca <- prcomp(x = param[, sapply(param, is.numeric)], scale. = TRUE)
summary(pca)

# Extract the 5th column of the list (which accounts for the first 5 PCs) and make it a dataframe

# The 1st-5th columns are the only PC's we really consider for this analysis, so no point looking beyond that because the influence of other PC values is going to be much less. We only use the first 2 PC's for plotting because it would be hard to visualize anything more than 2 dimensions 
pcascor <- as.data.frame(pca[[5]])
glimpse(pcascor)
pcaload <- as.data.frame(pca[[2]])
glimpse(pcaload)

# The pca loadings can be saved as a .csv file to review manually 
write.csv(pcaload, file = "RangewidePCAloadings_by_variable.csv")

# Add a column for call type (variant)
pcascor$variant <- sapply(1:nrow(param), function(x){
  strsplit(as.character(param$sound.files[x]), split = "_")[[1]][1]
})
View(pcascor)

# Create a data frame with the first 2 PC scores and the variants for plotting
pcascor_df <- data.frame(X = pcascor[, 1], Y = pcascor[, 2], variant = pcascor$variant)
hulls <- pcascor_df %>%
  group_by(variant) %>%
  slice(chull(X, Y))

# Create color palette
cols <- c("navajowhite4", "yellow", "darkorange", "tomato4") # "navyblue", "maroon2"

theme_AcousticSpace <- function(){
  theme(panel.background = element_rect(fill = "white"), plot.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "white"),
        panel.grid.minor = element_line(colour = "white"),
        axis.line = element_line(size = 0.8, colour = "black"),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 25),
        axis.text = element_text(size = 15))
}

# Generate plot
ggplot(pcascor_df, aes(x = X, y = Y)) +
    geom_point(aes(color = variant, fill = variant), shape = 21, size = 3) +
    geom_polygon(data = hulls, aes(x = X, y = Y, fill = variant, color = variant), alpha = 0.2, size = 0.2) +
    scale_colour_manual(values = cols) + scale_fill_manual(values = cols) +
    labs(x = "PC1", y = "PC2") +
    theme_AcousticSpace() +
    theme(legend.position = "right")
jpeg("PCA_plot.jpeg", units = NULL, width = 10, height = 11, res = 300)
  dev.off()

# Create data frame for ANOVA
adf <- data.frame(sels$Call_Type, pcascor$PC1, pcascor$PC2, pcascor$PC3, pcascor$PC3, pcascor$PC3)
names(adf) <- c("variant", "PC1", "PC2", "PC3", "PC4", "PC5")
View(adf)

# plot the mean PC1 by variant
a1plot <- boxplot(PC1 ~ variant, data = adf, xlab = "Variant", ylab = "PC1", main = "Mean PC1 Values by Variant")

a2plot <- boxplot(PC2 ~ variant, data = adf, xlab = "Variant", ylab = "PC2", main = "Mean PC1 Values by Variant")

# There are outliers in the residual plot that may imply the variances across groups are not homogeneous; we can use Levene's test to determine if they are significantly different
leveneTest(PC1 ~ variant, data = adf)
# Result: p-value is significant so variance between groups is significantly different

# Test data normality
shapiro.test(x = adf$PC1)
# result shows data is normal

# Need to use Welchs ANOVA since the data is normal but the variances are unequal
a1 <- oneway.test(PC1 ~ variant, adf)
a1
# data:  PC1 and variant
# F = 294.3, num df = 13.00, denom df = 299.41,
# p-value < 2.2e-16

a2 <- oneway.test(PC2 ~ variant, adf)
a2
# data:  PC2 and variant
# F = 53.292, num df = 13.00, denom df = 300.36,
# p-value < 2.2e-16

a3 <- oneway.test(PC3 ~ variant, adf)
a3
# data:  PC3 and variant
# F = 41.233, num df = 13.00, denom df = 302.52, p-value
# < 2.2e-16

a4 <- oneway.test(PC4 ~ variant, adf)
a4
# data:  PC4 and variant
# F = 41.233, num df = 13.00, denom df = 302.52, p-value
# < 2.2e-16

a5 <- oneway.test(PC5 ~ variant, adf)
a5
# data:  PC5 and variant
# F = 41.233, num df = 13.00, denom df = 302.52, p-value
# < 2.2e-16

TGH1 <- posthocTGH(adf$PC1, adf$variant, method = c("games-howell"), conf.level = 0.95, digits = 2, p.adjust = "none", formatPvalue = TRUE)
res1 <- print(TGH1)
View(res1)

TGH2 <- posthocTGH(adf$PC2, adf$variant, method = c("games-howell"), conf.level = 0.95, digits = 2, p.adjust = "none", formatPvalue = TRUE)
res2 <- print(TGH2)
View(res2)

TGH3 <- posthocTGH(adf$PC3, adf$variant, method = c("games-howell"), conf.level = 0.95, digits = 2, p.adjust = "none", formatPvalue = TRUE)
res3 <- print(TGH3)
View(res3)

TGH4 <- posthocTGH(adf$PC4, adf$variant, method = c("games-howell"), conf.level = 0.95, digits = 2, p.adjust = "none", formatPvalue = TRUE)
res4 <- print(a4)
write.csv(res4, file = "YNA_PCAposthoc_PC4.csv")

TGH5 <- posthocTGH(adf$PC5, adf$variant, method = c("games-howell"), conf.level = 0.95, digits = 2, p.adjust = "none", formatPvalue = TRUE)
res5 <- print(a5)
write.csv(res5, file = "YNA_PCAposthoc_PC5.csv")

```

